Concurrent Programming with CXL

### 2020-06-24 book/paper.tex
Made a pass over the deadlock chapter

\section*{Exercises}
\begin{problems}
\item Write a new version of the GPU allocator in Exercise~\ref{ex:gpu}.
In this version,
a process is allowed to allocate a set of GPUs and release a set of GPUs that it
has allocated.  Method \texttt{gpuAllocSet(n)} should block until $n$ GPUs are
available, but it should grant them as soon as they are available.
Method \texttt{gpuReleaseSet(s)} takes a set of GPU identifiers as argument.
A process does not have to return all the GPUs it allocated at once.
\item The specification in the previous question makes the solution unfair.
Explain why this is so.  Then change the specification and the solution so that
it is fair.
\end{problems}

@@

Preventing deadlock thus means preventing that one of these conditions occurs.
However, mutual exclusion is not easily prevented (although, for some resources it is
possible, as demonstrated in Chapter~\ref{ch:nonblocking}).
Havender~\cite{Havender68} proposed the following techniques that avoid the remaining
three conditions:

@@

Figure~\ref{fig:dinerscv} demonstrates how this might be done.
We use a single mutex for the diners, and, for each fork, a boolean
and a condition variable.  The boolean indicates if the fork has been
taken.
Each diner waits if either the left or right fork is already taken.
But which condition variable to wait on?
The code demonstrates an important technique to use when waiting for
multiple conditions.
\index{multiple conditions, waiting on}
The condition in the \texttt{while} statement is the condition that the
diner is waiting for to be \texttt{False}.  Within the \texttt{while} statement,
there is an \texttt{if} statement for each disjunct of that condition.
The code waits for either or both forks if necessary.  After that, it goes
back to the top of the \texttt{while} loop.

@@

This completes all the Havender methods.
There is, however, another approach, which is sometimes called
\emph{deadlock avoidance}.
\index{deadlock avoidance}
In the case of the Dining Philosophers, we want to avoid the situation where each
diner picks up a fork.  So if we can prevent more than four diners from starting to
eat at the same time, then we can avoid the conditions for deadlock from ever
happening.
Figure~\ref{fig:dinersavoid} demonstrates this concept.  It uses a semaphore to
restrict the number of diners at any time to four.

This concept can be generalized using something called the
Banker's Algorithm~\cite{EWD108}, but it is outside the scope of this book.
The problem with these kinds of schemes is that one needs to know ahead of time
what the maximum number of resources is that each process wants to allocate,
making them generally quite impractical.

@@

\section*{Exercises}
\begin{problems}
\item Figure~\ref{fig:bank} shows an implementation of a bank with various
accounts and transfers between those accounts.
Unfortunately, running the test reveals that it sometimes leaves unterminated
processes.  Can you fix the problem?
\item Add a method \texttt{total()} to the solution of the previous question
that computes the total over all balances.
It needs to obtain a lock on all accounts.  Make sure that
it cannot cause deadlock.
\item Use method \texttt{total()} to ensure that the sum total of the bank
accounts at the beginning of the test is the same as at the end.
\end{problems}

>>>>

\section*{Exercises}
\begin{problems}
\item Implement semaphores using Mesa condition variables.
\item Write a new version of the GPU allocator in Exercise~\ref{ex:gpu}
using Mesa condition variables.
In this version,
a process is allowed to allocate a set of GPUs and release a set of GPUs that it
has allocated.  Method \texttt{gpuAllocSet(n)} should block until $n$ GPUs are
available, but it should grant them as soon as they are available.
Method \texttt{gpuReleaseSet(s)} takes a set of GPU identifiers as argument.
A process does not have to return all the GPUs it allocated at once.
\item The specification in the previous question makes the solution unfair.
Explain why this is so.  Then change the specification and the solution so that
it is fair.
\end{problems}

@@

Preventing deadlock thus means preventing that one of these conditions occurs.
However, mutual exclusion is not easily prevented (although, for some resources it is
possible, as demonstrated in Chapter~\ref{ch:nonblocking}).
Havender proposed the following techniques that avoid the remaining
three conditions~\cite{Havender68}:

@@

Figure~\ref{fig:dinerscv} demonstrates how this might be done.
We use a single mutex for the diners, and, for each fork, a boolean
and a condition variable.  The boolean indicates if the fork has been
taken.
Each diner waits if either the left or right fork is already taken.
But which condition variable to wait on?
The code demonstrates an important technique to use when waiting for
multiple conditions.
\index{multiple conditions, waiting on}
The condition in the \texttt{while} statement is the negation of the
condition that the diner is waiting and consists of two disjuncts.
Within the \texttt{while} statement,
there is an \texttt{if} statement for each disjunct.
The code waits for either or both forks if necessary.  After that, it goes
back to the top of the \texttt{while} loop.

@@

This completes all the Havender methods.
There is, however, another approach, which is sometimes called deadlock
\emph{avoidance}
\index{deadlock avoidance}
instead of deadlock \emph{prevention}.
In the case of the Dining Philosophers, we want to avoid the situation where each
diner picks up a fork.  So if we can prevent more than four diners from starting to
eat at the same time, then we can avoid the conditions for deadlock from ever
happening.
Figure~\ref{fig:dinersavoid} demonstrates this concept.  It uses a semaphore to
restrict the number of diners at any time to four.

This concept can be generalized using something called the
Banker's Algorithm~\cite{EWD108}, but it is outside the scope of this book.
The problem with these kinds of schemes is that one needs to know ahead of time
the set of processes and what the maximum number of resources is that each process
wants to allocate, making them generally quite impractical.

@@

\section*{Exercises}
\begin{problems}
\item Figure~\ref{fig:bank} shows an implementation of a bank with various
accounts and transfers between those accounts.
Unfortunately, running the test reveals that it sometimes leaves unterminated
processes.  Can you fix the problem?
\item Add a method \texttt{total()} to the solution of the previous question
that computes the total over all balances.
It needs to obtain a lock on all accounts.  Make sure that
it cannot cause deadlock.
\item Implement an extra process that checks, once, if the total of
the balances is the same as that at the beginning (using an assertion).
Is once enough to determine that no money gets lost or create, or should it do
so continually?
\end{problems}
